---
title: "Kommuneendringer"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I often find myself working with data that deals with Norwegian municipalities (kommuner) and counties (fylker), either in individual level data (a persons residence, for example) or on data on the municipal level itself. The standard way of working with municipalities is to use the municipality number, which is a four digit number, where the two first digits indicate the county number. In the recent years  (2018 - 2024) there has been large reforms of the municipal and county structure, causing a lot of frustration when working with municipal data, especially if the data spans several years. I often find myself just wanting to translate the code for a municipality from one point in time to another point in time, while taking into account mergers, splits, name changes, county changes etc. The goal of the `kommuneendringer` R package is to do just that. 

This package is *opinionated*, in the sense that it works in ways that suits my needs, which might be contrary to official guidelines for how to deal with changes in the municipality structure. Please take a good look at the examples and explanations below to get an understanding of how it works.


## Installation

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("opisthokonta/kommuneendringer")
```

## Features and limitatons

The main functionality of this package is the `translate_knr` function. This can translate the four-digit municipal code between different time point starting from January 1st 1977. It is updated with all changes up until January 1st 2024.

It currently only supports translating municipality codes. I will add functionality for translating county codes at some point in the future. 

Special municipal codes such as those for Svalbard, off-shore installations , etc, are not supported. 

In 2024 some mergers between counties and municipalities from 2020 were reversed. I hope to add support for taking this into account in the future. 


## Examples

First load the package:
```{r}
library(kommuneendringer)
```

### Example 1: Changing county

A simple example is the municipality Halden. It hasn't merged with any other municipality recently, but changed county, and therefore its code, in 2020 (from Østfold to Viken). The code was changed again in 2024, when Viken was split and the county of Østfold was re-established. The new Østfold county got a new county number and Halden's new code was therefore not the same as before 2020.

```{r}
# Change code from 0101 to 3001 on January 1st 2020.
translate_knr('0101', from_date = '2019-01-01', to_date = '2020-05-17')
```
Lets see what it changed to in 2024:
```{r}
translate_knr('3001', from_date = '2020-01-01', to_date = '2024-05-17')
```
It is also possible to make longer jumps, say if you have some Halden data from 2008, and want to merge it with data from 2024:

```{r}
translate_knr('0101', from_date = '2008-01-01', to_date = '2024-05-17')
```
It is also possible to go backwards in time:

```{r}
translate_knr('3101', from_date = '2024-01-01', to_date = '2015-05-17')
```

### Example 2: Merged municipalities

An example of municipalities that merged are Søgne, Songdalen and Kristiansand, which merged in 2020. When municipalities merge, the `translate_knr` function will, if possible, give you the code for the merged municipality. You can then perhaps use this to aggregate data from the historical municipalities as if they are the new municipality. 


```{r}
# Søgne's code was 1018.
translate_knr('1018', from_date = '2018-01-01', to_date = '2023-05-17')
```
```{r}
# Songdalen's code was 1017
translate_knr('1017', from_date = '2018-01-01', to_date = '2023-05-17')
```
```{r}
# Kristiansand's code was 1001
translate_knr('1001', from_date = '2018-01-01', to_date = '2023-05-17')
```
Let's take a look what happens if you try to translate the code backwards in time:

```{r}
translate_knr('4204', from_date = '2023-01-01', to_date = '2015-05-17')
```
The function will return NA and a warning, since there is no unique predecessor municipality. The municipality with code 4204 in 2023 can correspond to any of the three predecessors (1001, 1017 and 1018) in 2015, and without any other information, we can not know which one is the correct one.


### Example 3: Split municipality

Tysfjord municipality (code 1850) was in 2020 split in two, and the two parts merged with other municipalities. There is no unique successor municipality and it is therefore not possible to translate the code from before to after January 1st 2020.

```{r}
translate_knr('1850', from_date = '2019-01-01', to_date = '2020-05-17')
```



## How it works

This pacakge is based on data from Statistics Norway (SSB). From this data a directed graph (or network) is created, where a municipality is represented by a node. When the name or code of the municipality change, the municipality is represented by a new node, and there is a directed edge pointing from the old municipality to the new. New nodes are created when:

* Code change
* Name change (including when the municipality get official sami and kven names)
* Merger, even when the new municipality keep name and code from predecessors.


When mergers occur, two (or more) nodes representing the old municipalities have edges pointing to the new merged municipality. Each node have attributes that marks the start date and end date of the municipality. This way the history of a municipality can be followed and the correct successor municipalities can be found after mergers.



## Regarding changes in the municipal codes

So dealing with changes in the municipal codes are frustrating for a number of reasons. Typically, a municipality change code  

* When municipalities or counties are merged. The merged municipality will usually get a new code, even if it keeps the name of one of its predecessors. But there are exceptions to this rule, which is documented below.

* When municipalities or counties split. The new municipalities or counties will each get a new code. 

* When municipalities change county. Since the first two digits of the municipal code indicate the county, the code will change correspondingly. This also happens when counties merge or split. 


The code will not change 

* When the municipality changes name. This includes when a municipality get an additional official name in one of the minority langues (sami or kven). 

Code and name changes always takes place on January 1st. 

### Exceptional cases

This section is mostly a note to myself, to keep track of those instances that need particular attention to get right.

Usually when municipalities are merged, they get a new code, but this has not always happened:

* When 0716 Våle and 0718 Ramnes were merged to become Re in 2002, the new municipality kept Våle's code 0716.
* When 1804 Bodø and 1842 Skjerstad were merged in 2005, the kept the name and code for 1804 Bodø.
* When 1141 Finnøy and 1142 Rennesøy were merged with 1103 Stavanger in 2020, the new municipality kept the name and code for 1103 Stavanger.
* When 5030 Klæbu was merged with 5001 Trondheim in 2020, the new municipality kept the name and code for 5001 Trondheim.


#### Municipalities that split

Usually, when there are changes in the municipal structure, it is because two or more municipalities merge, but there are exceptions to this, when sometimes municipalities split:

* In 2020, the municipality 1850 Tysfjord was split in two. One part merged with 1805 Narvik to become 1806 Narvik, and the other part merged with 1849 Hammarøy to become 1875 Hammarøy.
*  In 2020, the municipality 5012 Snillfjord was split in 3. One part was merged with 3 other muncipalities to become 5059 Orkland. The second part was merged with two other municipalities to become 5055 Heim. The third part was merged with 5013 Hitra to become 5056 Hitra.
* In 2020, the municipalities 1504 Ålesund and 1534 Haram merged to become 1507 Ålesund. In 2024 they reversed the merger, and split into 1508 Ålesund and 1580 Haram.
* In 2017, the municipality 0720 Stokke was merged with 0706 Sandefjord (together with some other municipalities) to become 0710 Sandefjord, but part of it was merged with 0704 Tønsberg. Hence it is considered a split.
* In 1992, part of 0412  Ringsaker was given to 0403 Hamar, and this is considered a split. 

Splits not taken account of in data:
* In 2002, 1514 Sande gave part to 1511 Vanylven, no change in code or names occured.
* In 1992, small part of 0114 Varteig was given to Rakkestad, but the main part of the municiaplity was merged with Sandefjord.

There might be other smaller border adjustments that are not in the data.


#### Municipalities that share names
There are some municipalities that share the same name. Conventionally the name of the county is added in parenthesis to distinguish them. 

* Våler (Innlandet, before 2020: Hedmark), and Våler (Østfold, in 2020-2023: Viken).
* Herøy (Møre og Romsdal) and Herøy (Nordland).
* Nes (Akershus) Nes (Buskerud). Nes in Buskerud changed name to Nesbyen in 2020 since both municipalities then ended up in the same county (Viken) when Akershus, Buskerud and Østfold merged in 2020. After Viken split in 2024, Nesbyen kept its new name.
* Os (Hedmark) and Os (Hordaland). Os in Hordaland was merged with another municipality in 2020 and got a new name. 
* Bø (Nordland), Bø (Telemark). Bø in Telemark merged with another municipality in 2020 and got a new name.
* Sande (Vestfold) and Sande (Møre og Romsdal). Sande in Vestfold merged with another municipality in 2020 and got a new name.



Samiske navn på karasjok og nesseby, dato..

## Version history
* 0.1.3 (March 2025) - Some changes to in the underlying code to build the graph object from SSB data. This should make it easier to update the pacakge with more historical data and new changes in the future. Should be no changes for the user. 
* 0.1.2 (January 2025) - Updated data from SSB (data was updated in September 2024). Results will sometimes be different from what they were in version 0.1.1, since some historical municipality mergers/splits/changes have been added to the data.
* 0.1.1 - Faster code conversion. The function now re-uses results when input is duplicated, instead of traversing the graph multiple times.
* 0.1.0 (July 2024) - Initial version.


## Links

The data is sourced from Statistics Norway (Statistisk Sentralbyrå, SSB). Here are some relevant links.

* [Alle endringer i de regionale inndelingene](https://www.ssb.no/metadata/alle-endringer-i-de-regionale-inndelingene) (in Norwegian)
* [Classification of municipalities](https://www.ssb.no/en/klass/klassifikasjoner/131/)

Wikipedia has a relevant article:

* [Wikipedia:Norske kommunenummer](https://no.wikipedia.org/wiki/Norske_kommunenummer) (in Norwegian)







